---
title: "Workshop 2"
author: "zwx"
format: html
editor: visual
---

## File I/O

The path below is relative to this Quarto document. The first column in `Sonar.csv` contains row numbers, so `row.names = 1` prevents it from being read as an additional variable.

### a) Read and inspect the data

```{r}
sonar <- read.csv("../Datasets/Sonar.csv", row.names = 1)

# Confirm that sonar is a data frame
is.data.frame(sonar)

# Investigate the first six rows
head(sonar)
```

`is.data.frame(sonar)` returns `TRUE`.

### b) Read `Class` as a factor

```{r}
sonar <- read.csv(
  "../Datasets/Sonar.csv",
  row.names = 1,
  # Include the original row-number column in the 62 CSV columns.
  colClasses = c("character", rep("numeric", 60), "factor")
)

# Check the type of Class
is.factor(sonar$Class)
class(sonar$Class)
```

The checks confirm that `Class` is a factor.

### c) Number of levels in `Class`

```{r}
nlevels(sonar$Class)
levels(sonar$Class)
```

There are **2 levels**, `M` and `R`.

### d) Work with `V1` as a vector

```{r}
v1 <- sonar$V1

# Confirm that v1 is a vector and find its length
is.vector(v1)
length(v1)

# Output elements 5 through 10 (inclusive)
v1[5:10]

# Append 1.5 to the end of the vector
v1 <- c(v1, 1.5)
tail(v1)
length(v1)
```

Originally, `v1` contains **208 elements**. After appending `1.5`, it contains **209 elements**.

### e) Convert to a matrix and remove `Class`

```{r}
sonar_matrix <- as.matrix(sonar)
is.matrix(sonar_matrix)

# Remove the Class column while retaining the matrix structure
sonar_matrix <- sonar_matrix[, colnames(sonar_matrix) != "Class", drop = FALSE]

# Confirm the result
is.matrix(sonar_matrix)
dim(sonar_matrix)
colnames(sonar_matrix)
```

After removing `Class`, `sonar_matrix` has **208 rows and 60 columns**.

## Numerical Summary

### a) Quartiles and the shape of the `V1` distribution

The `quantile()` function can return the first quartile, median and third quartile in a single call.

```{r}
quantile(sonar$V1, probs = c(0.25, 0.50, 0.75))

# Compare the mean with the median
mean(sonar$V1)
median(sonar$V1)
```

The first quartile is **0.01335**, the median is **0.02280**, and the third quartile is **0.03555**. The mean (**0.02916**) is greater than the median (**0.02280**), suggesting that the distribution of `V1` is **right-skewed (positively skewed)**. In other words, some relatively large values extend the right tail and pull the mean upward.

### b) Maximum, minimum, standard deviation and mean of `V1`

```{r}
max(sonar$V1)
min(sonar$V1)
sd(sonar$V1)
mean(sonar$V1)
```

The maximum is **0.13710**, the minimum is **0.00150**, the standard deviation is approximately **0.02299**, and the mean is approximately **0.02916**.

### c) Median of `V1` for each `Class` using tidyverse

```{r}
library(tidyverse)

sonar |>
  group_by(Class) |>
  summarise(median_V1 = median(V1), .groups = "drop")
```

The median of `V1` is **0.0269** for class `M` and **0.0201** for class `R`.

## Graphical Summary

The following plots use `ggplot2`, which is included in the tidyverse package loaded above.

### Boxplot of `V1` against `Class`

```{r}
ggplot(sonar, aes(x = Class, y = V1)) +
  geom_boxplot() +
  labs(
    title = "Distribution of V1 by Class",
    x = "Class",
    y = "V1"
  ) +
  theme_minimal()
```

### Scatterplot of `V2` against `V1`

```{r}
ggplot(sonar, aes(x = V1, y = V2)) +
  geom_point() +
  labs(
    title = "V2 against V1",
    x = "V1",
    y = "V2"
  ) +
  theme_minimal()
```

## Write Data to File

First, filter the data to retain only observations in class `M`. Then use `write.csv()` to save the filtered data as `mine.csv`.

```{r}
mine <- sonar |>
  filter(Class == "M")

write.csv(mine, "../Datasets/mine.csv", row.names = FALSE)

# Check the filtered data
nrow(mine)
unique(mine$Class)
```

The new dataset contains **111 observations**, all belonging to class `M`. Setting `row.names = FALSE` prevents R from adding a separate row-number column to the CSV file.
