---
title: "Workshop 1"
author: "Wenxuan Zhao"
date: "2026-08-06"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
```

## File Input/Output

### a) Read and inspect the data

Read `Cereal.csv` into R, display its first six observations, and check the class of the resulting object.

```{r}
cereal <- read.csv("../Datasets/Cereal.csv")

head(cereal)
class(cereal)
```

The data are stored as a data frame, with one row for each cereal and one column for each variable.

### b) Examine the dimensions

```{r}
colnames(cereal)
nrow(cereal)
dim(cereal)
```

`colnames()` returns the variable names, `nrow()` gives the number of observations, and `dim()` reports both the number of rows and columns.

### c) Extract the calorie variable

```{r}
Cal <- cereal$calories

head(Cal)
```

### d) Retain the first 10 observations

```{r}
cereal_first_10 <- cereal[1:10, ]

cereal_first_10
```

Using a new object preserves the complete `cereal` data frame for the remaining questions.

### e) Select Kellogg's cereals

```{r}
Kelloggs <- cereal_first_10[cereal_first_10$mfr == "K", ]

Kelloggs
```

The manufacturer code `K` represents Kellogg's.

## Factors

### a) Read character variables as factors

```{r}
cereal_with_factors <- read.csv(
  "../Datasets/Cereal.csv",
  stringsAsFactors = TRUE
)

str(cereal_with_factors)
```

### b) Inspect factor levels

```{r}
levels(cereal_with_factors$mfr)
nlevels(cereal_with_factors$type)
```

`levels()` lists the manufacturer categories, while `nlevels()` returns the number of cereal-type categories.

## Vectors

### a) Create a calorie vector

```{r}
cereal_calories <- cereal_with_factors$calories
```

### b) Find the vector length

```{r}
length(cereal_calories)
```

### c) Extract elements 5 through 10

```{r}
cereal_calories[5:10]
```

### d) Append a value

```{r}
cereal_calories <- c(cereal_calories, 100)

tail(cereal_calories)
length(cereal_calories)
```

## Matrices

### a) Convert the data frame to a matrix

```{r}
cereal_matrix <- as.matrix(cereal)

is.matrix(cereal_matrix)
is.character(cereal_matrix)
```

The matrix is a character matrix because a matrix can contain only one data type, and the original data include character columns.

### b) Create a numeric matrix

Remove the non-numeric variables before converting the data frame to a matrix.

```{r}
cereal_numeric_matrix <- cereal |>
  select(-mfr, -name, -type) |>
  as.matrix()

is.matrix(cereal_numeric_matrix)
is.numeric(cereal_numeric_matrix)
```

## Numerical Summary

### a) Summarise sodium content

```{r}
summary(cereal$sodium)
```

### b) Calculate individual summary statistics

```{r}
max(cereal$sodium)
min(cereal$sodium)
sd(cereal$sodium)
mean(cereal$sodium)
```

### c) Calculate mean sodium by manufacturer

```{r}
cereal |>
  group_by(mfr) |>
  summarise(
    mean_sodium = mean(sodium),
    .groups = "drop"
  )
```

## Graphical Summary

### a) Compare sodium across manufacturers

```{r}
boxplot(
  sodium ~ mfr,
  data = cereal,
  xlab = "Manufacturer",
  ylab = "Sodium (mg)",
  main = "Sodium Content by Manufacturer",
  col = "lightblue"
)
```

### b) Examine calories against sodium

```{r}
plot(
  cereal$sodium,
  cereal$calories,
  xlab = "Sodium (mg)",
  ylab = "Calories",
  main = "Calories Against Sodium",
  pch = 19,
  col = "steelblue"
)
```

## Write Data to a File

### a) Export all Kellogg's cereals

Filter the complete dataset for Kellogg's cereals and save the result as a CSV file. Setting `row.names = FALSE` prevents R from adding row numbers as an extra column.

```{r}
Kelloggs <- cereal |>
  filter(mfr == "K")

write.csv(
  Kelloggs,
  "../Datasets/kelloggs.csv",
  row.names = FALSE
)
```
